---
title: Barcode analysis for select groups of mimetic populations
date: "`r Sys.Date()`"
---

# Background

Investigate whether the distribution of barcodes in the two major groups
supports the hypothesis of separate progenitors for them.

```{r setup, message = FALSE}
library(tidyverse)
library(scater)
library(patchwork)
devtools::load_all(file.path(ROOT, "R"))
FIGS <- file.path(ROOT, "analysis", "output", "figures", "sc")
dir.create(FIGS, recursive = TRUE, showWarnings = FALSE)

# Load integrated data with identified mimetics
sce <- readRDS(file.path(ROOT, "pipeline", "integrated_universe", "sce_integrated_mimetics.rds"))
```


```{r}
sce <- sce[, sce$genotype == "WT" & sce$timepoint %in% c("embryo", "newborn", "4week") & sce$barcoded == TRUE]
metadata <- as.data.frame(colData(sce)) %>%
    select(sample, genotype, timepoint) %>%
    distinct()
rownames(metadata) <- NULL
```

Metadata of samples still contributing to this analysis:

```{r}
metadata
```


```{r}
# Background frequencies of barcodes/scars
background_bulk <- read.csv(file.path(ROOT, "data", "sc", "background.csv"))

# Group by population or mimetic label
# Prefer mimetic labels. Do not double count cells.
bc_counts_all <- as_tibble(colData(sce)) %>%
    mutate(
        group = case_when(
            signature == "Aire-stage" ~ "Aire-stage",
            signature != "no_signature" ~ "mimetic",
            label %in% c("early_progenitor", "postnatal_progenitor") ~ label
        )
    ) %>%
    filter(!is.na(group)) %>%
    group_by(sample, bcid, group) %>%
    summarize(count = n()) %>%
    ungroup() %>%
    mutate(group = forcats::fct_relevel(group, c("early_progenitor", "postnatal_progenitor", "Aire-stage"))) %>%
    group_by(sample, bcid) %>%
    mutate(total_bc_count = sum(count)) %>%
    ungroup() %>%
    filter(
        # IMPORTANT!
        # Filtering of uninformative barcodes
        !(is.na(bcid)),
        bcid != "WT"
    ) %>%
    complete(nesting(sample, bcid), group, fill = list(count = 0)) %>%
    group_by(sample, group) %>%
    mutate(N = sum(count)) # number of cells with barcodes in population
# Store final used signature labels
all_groups <- sort(unique(bc_counts_all$group))
```


# Barcode Sampling Probabilities

This shows the p value (probability of sampling this many or more of a barcode
within a cell population) for all barcodes in all populations. It is analogous
to the analysis conducted by DG. However, barcodes for all mice were plotted in
a single heatmap.

Only barcodes with an adjusted p value of ≤ 0.05 in at least one population are
*shown*, but they do still contribute to p value correction. This reduces
clutter on the heatmap but is still less strict than what is deemed
"informative" in Nusser et al. 2022 (Methods, p < 0.001).

P values for individual barcodes carry information about a barcode appearing by
chance in a population, given its sample size (number of cells with barcodes)
and the rarity of the respective barcode.


```{r fig_4w_barcode_pvals, fig.height = 2, fig.width = 6}
# p values for individual barcodes in populations, analogous to DG analysis
bc_p_vals <- bc_counts_all %>%
    # filter(label != "unassigned" & label != "no_signature") %>%
    # mutate(label = droplevels(label)) %>%
    mutate(
        sfreq = get_background_frequency(bcid, background_bulk),
        # Make barcode identifiers unique over samples to show all in one heatmap
        bcid_sample = paste(sample, bcid, sep = ".bc")
    ) %>%
    group_by(sample, group, bcid, bcid_sample) %>%
    summarize(
        count = count,
        sfreq = sfreq,
        N = N,
        p = binom.test(count, n = max(1, N), p = sfreq, alternative = "g")$p.value,
        p = ifelse(N > 0, p, NA)
    ) %>%
    # Correct p value per sample
    group_by(sample) %>%
    mutate(p.adj = barcode_p_adj(p, count)) %>%
    # Filter completely uninteresting barcodes to reduce clutter
    group_by(sample, bcid) %>%
    filter(any(p.adj <= 0.05)) %>%
    # add metadata
    left_join(metadata) %>%
    mutate(log.p.adj = -log10(p.adj + 1e-10))

# Determine plot order by clustering
bcid_tab <-  with(bc_p_vals, xtabs(log.p.adj ~ bcid_sample + group))
bcid_ord <- hclust(dist(bcid_tab))$order
bc_p_vals$bcid_sample <- factor(bc_p_vals$bcid_sample, levels = rownames(bcid_tab)[bcid_ord])

p_crispr_bc <- ggplot(data = NULL, aes(bcid_sample, group, fill = log.p.adj)) +
    geom_tile() +
    scale_y_discrete(limits = rev, expand = c(0, 0), labels = LBL_SIGNATURE) +
    scale_x_discrete(
        expand = c(0, 0),
        labels = function(x) {
            x <- gsub(".+[.]bc", "", x)
            x <- gsub("_", "/", x)
            x
        }
    ) +
    scale_fill_viridis_c(breaks = c(0, 5, 10), labels = c(0, 5, "> 10")) +
    facet_grid(
        ~sample,
        scales = "free_x", space = "free",
        labeller = as_labeller(c(
            "embryo" = "E16.5",
            "newborn" = "P0",
            "W4BCFEM1" = "P28 (#1)",
            "W4BCFEM2" = "P28 (#2)",
            "W4BCMALE" = "P28 (#3)"
        ))
    ) +
    labs(
        x = "CRISPR barcode",
        y = NULL,
        fill = quote(-log[10]*"(adj. p)")
    ) +
    guides(fill = guide_colorbar(title.position = "top")) +
    theme(
        legend.direction = "horizontal",
        legend.key.width = unit(5, "mm"),
        legend.key.height = unit(2, "mm"),
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, margin = unit(c(1, 1, 1, 1), "pt"))
    ) +
    geom_hline(yintercept = seq(0.5, 10.5, by = 1), color = "gray90", linewidth = 0.1)
row1 <- c("embryo", "newborn")
p_crispr_bc1 <- (p_crispr_bc %+% filter(bc_p_vals, sample %in% row1)) +
    theme(
        axis.title.x = element_blank(),
        legend.box.margin = unit(c(0, 0, 0, 10), "mm")
    )
p_crispr_bc2 <- (p_crispr_bc %+% filter(bc_p_vals, !(sample %in% row1))) +
    theme(legend.position = "none")
p_crispr_bc <- (
    (
        (p_crispr_bc1 + guide_area()) + plot_layout(widths = c(3, 1))
    ) /
        (p_crispr_bc2)
    )
p_crispr_bc

write.table(
    bc_p_vals[c("sample", "timepoint", "group", "bcid", "log.p.adj")],
    file.path(FIGS, "tbl_barcode_pvals.tsv"),
    sep = "\t",
    row.names = FALSE
)
fig_pdf(FIGS, p_crispr_bc)
```
