---
title: Integration of data from different time points and filtering of gene tables
date: "`r Sys.Date()`"
params:
    universe: TRUE
    # if TRUE, keep only genes that are present in ALL samples
    # if FALSE, keep genes present in ANY sample and set missing counts to 0
    sce_out: NULL
---

```{r setup, message = FALSE, results = "hide"}
library(tidyverse)
library(scran)
library(scater)
library(batchelor)
devtools::load_all(file.path(ROOT, "R"))  # basically library(mimetics)

universe <- params$universe
stopifnot(!is.null(params$sce_out))
```


# Integrate data from different time points

To facilitate uniform visualization of mimetic cell populations, unify all data
sets in one consolidated object.

Unify the following samples:


```{r}
sample_names <- c(
    "embryo",
    "newborn",
    "W4BCFEM1",
    "W4BCFEM2",
    "W4BCMALE",
    "W4WTMALE"
)
sample_names
```

```{r load-data, warning = FALSE}
sce_paths <- file.path(ROOT, "pipeline", sprintf("sce_%s.rds", sample_names))
sces <- lapply(sce_paths, readRDS)
sce <- do.call(combineCols, unname(sces))
```


```{r universe-yes, eval = universe, results = "asis"}
# keep genes present in ALL samples
gene_universe <- lapply(sces, rownames)
gene_universe <- Reduce(intersect, gene_universe)
sce <- sce[gene_universe, ]
```

```{r universe-no, eval = !universe}
# keep genes present in ANY sample and fill NAs with 0
gene_missing <- is.na(assay(sce, "counts"))
assay(sce, "counts")[gene_missing] <- 0
assay(sce, "logcounts")[gene_missing] <- 0
```

There are **`r nrow(sce)` genes** in the data.

```{r}
# Realize delayed matrices
assays(sce) <- lapply(assays(sce), as.matrix)
```


# UMAP

Run a new UMAP dimensionality reduction for all data together.


## Before integration

Prior to data integration, the effect of different samples dominates for UMAP
dimensionality reduction. The different populations (as defined by the previous
RaceID analysis) do not overlap for different samples.

```{r umap-before-integration}
set.seed(123456)
sce <- runUMAP(sce, min_dist = 0.5, metric = "euclidean", name = "UMAP_all", ntop = 2000, pca = 100)
plotReducedDim(sce, dimred = "UMAP_all", color_by = "sample")
plotReducedDim(sce, dimred = "UMAP_all", color_by = "label")
```


## After integration

After integration, the labeled populations overlap across samples.

```{r integration, warning = FALSE}
set.seed(123456)

# First integrate within time points, then between time points
# Set timepoint levels to ensure correct merge order
colData(sce)$timepoint <- factor(
    colData(sce)$timepoint,
    levels = c(
        "embryo",
        "newborn",
        "4week"
    )
)
metadata(sce)[["get_merge.order"]] <- function(sce) {
    with(droplevels(unique(colData(sce)[, c("sample", "timepoint")])),
        split(sample, timepoint)
    )
}

# Set label levels for plotting order
colData(sce)$label <- factor(colData(sce)$label)
colData(sce)$label <- forcats::fct_relevel(colData(sce)$label,
    "early_progenitor",
    "postnatal_progenitor",
    "mTEC",
    "cTEC",
    "cTEC_nurse"
)


run_fastMNN <- function(sce) {
    sce %>%
        batchelor::multiBatchNorm(., batch = .[["sample"]]) %>%
        batchelor::fastMNN(
            batch = .[["sample"]],
            merge.order = metadata(.)[["get_merge.order"]](.),
            subset.row = getTopHVGs(modelGeneVar(., block = .[["sample"]])),
            BPPARAM = BiocParallel::MulticoreParam()
        )
}

sce_corr <- run_fastMNN(sce)

sce_4w <- sce[, sce$timepoint == "4week"]
sce_corr_4w <- run_fastMNN(sce_4w)
mat_corr_4w <- reducedDim(sce_corr_4w, "corrected")
# Expand week4 correction matrix to allow adding to full data set
mat_corr_4w_exp <- matrix(
    nrow = nrow(reducedDim(sce_corr, "corrected")),
    ncol = ncol(mat_corr_4w)
)
rownames(mat_corr_4w_exp) <- rownames(reducedDim(sce_corr, "corrected"))
mat_corr_4w_exp[rownames(mat_corr_4w),] <- mat_corr_4w

sce_corr <- runUMAP(sce_corr,
    min_dist = 0.5,
    dimred = "corrected",
    name = "UMAP_int",
    pca = 100)
reducedDim(sce, "UMAP_int") <- reducedDim(sce_corr, "UMAP_int")
reducedDim(sce, "fastMNN_all") <- reducedDim(sce_corr, "corrected")
reducedDim(sce, "fastMNN_4w") <- mat_corr_4w_exp
plotReducedDim(sce, "UMAP_int", color_by = "label")
```

```{r fig.height = 6}
ggcells(sce, aes(UMAP_int.1, UMAP_int.2, color = label)) +
    geom_point(size = 0.1) +
    facet_grid(genotype ~ timepoint + barcoded, labeller = label_both) +
    mimetics:::theme_umap
```


```{r save}
# Save integrated sce object, including metadata
saveRDS(sce, file.path(ROOT, params$sce_out))
```
