---
title: Filtering and consolidation of marker gene tables
date: "`r Sys.Date()`"
---

```{r setup, message = FALSE}
library(tidyverse)
MARKERS_DIR <- file.path(ROOT, "data", "markers")
```

Michelson et al. have provided lists of differentially regulated genes in their
clusters of mimetic populations vs. other cells (Supplementary Table S2). Here,
these signature are used to identify corresponding mimetic cells in our dataset
(Nusser et al. Nature 2022).


# Filtering of Table S2

```{r s2-filters}
thresh_padj <- 1e-2
thresh_log2fc <- 1
thresh_pct1 <- 0.1
```

The differentially expressed genes for each mimetic population are filtered:

* Adjusted p value <= `r thresh_padj`
* log2 fold change >= `r thresh_log2fc`
* Percent of cells expressing gene in cluster >= `r thresh_pct1 * 100`

```{r markers-michelson}
# Utilizing data from Michelson et al., Table S2
mimetic_markers <- read.csv(file.path(MARKERS_DIR, "michelson_tbls2.csv")) %>%
    filter(p.adj <= thresh_padj, avg_log2fc >= thresh_log2fc, pct1 >= thresh_pct1)
```

We also fetch data for signatures of TEC populations from Nusser et al., Tables
S1 and S2.

```{r markers-nusser}
# Data from Nusser et al., Table S1 and S2
tec_markers <- read.csv(file.path(MARKERS_DIR, "nusser_progenitor_signatures.csv"))
```


Combine TEC and mimetic markers and save as list.

```{r combine}
markers <- bind_rows(mimetic_markers, tec_markers) %>%
    group_by(signature) %>%
    summarize(gene = list(gene)) %>%
    deframe()
saveRDS(markers, file.path(ROOT, "pipeline", "markers.rds"))
```


Based on subsequent analysis (see identify_mimetics.Rmd), collapse similar
signatures and restrict to mimetic populations of interest.

```{r collapse}
mimetic_markers_collapsed <- mimetic_markers %>%
    mutate(
        signature = ifelse(signature %in% c("Tuft1", "Tuft2"), "Tuft", signature),
        signature = ifelse(signature %in% c("Skin_basal", "Skin_keratinized"), "Skin", signature)
    )
markers_collapsed <- bind_rows(mimetic_markers_collapsed, tec_markers) %>%
    group_by(signature) %>%
    summarize(gene = list(unique(gene))) %>%
    deframe()
saveRDS(markers_collapsed, file.path(ROOT, "pipeline", "markers_collapsed.rds"))


# Restrict to mimetic signatures of interest
signatures <- c("Ciliated", "Goblet", "Ionocyte", "Neuroendocrine", "Muscle", "Lung_basal", "EnteroHepato", "Pancreatic", "Microfold", "Skin", "Tuft", "Aire-stage")
markers_mimetics_collapsed <- markers_collapsed[signatures]
saveRDS(markers_mimetics_collapsed, file.path(ROOT, "pipeline", "markers_mimetics_collapsed.rds"))

# Store a non-redundant set of genes that are specific to only one signature
sigs_per_gene <- table(unlist(markers_mimetics_collapsed))
unique_genes <- names(sigs_per_gene[sigs_per_gene == 1])
markers_mimetics_collapsed_unique <- lapply(markers_mimetics_collapsed, function(x) {
    intersect(x, unique_genes)
})
saveRDS(markers_mimetics_collapsed_unique, file.path(ROOT, "pipeline", "markers_mimetics_collapsed_unique.rds"))
```


Also store PanglaoDB markers for control purposes.

```{r panglao}
markers_panglao <- read.csv(file.path(MARKERS_DIR, "PanglaoDB_markers_27_Mar_2020.tsv"), sep = "\t") %>%
    filter(
        grepl("Mm", species),
        gene.type == "protein-coding gene"
    ) %>%
    mutate(
        signature = paste0("Panglao:", organ, ".", cell.type),
        gene = stringr::str_to_title(official.gene.symbol)
    ) %>%
    group_by(signature) %>%
    summarize(gene = list(unique(gene))) %>%
    deframe()
saveRDS(markers_panglao, file.path(ROOT, "pipeline", "markers_panglao.rds"))
```


And Tabula Muris signatures.

```{r tmuris}
read_tm_markers <- function(tissue, method) {
    basedir <- file.path(MARKERS_DIR, "tabula_muris_markers")
    f_classes <- file.path(basedir, paste(method, tissue, "cell_ontology_class", "classes.csv", sep = "_"))
    f_markers <- file.path(basedir, paste(method, tissue, "cell_ontology_class", "markers.csv", sep = "_"))
    classes <- read.csv(f_classes)
    classes <- classes[, -1, drop = FALSE]
    colnames(classes)[1] <- "cell_type"
    classes$cluster <- 0:(nrow(classes) - 1)
    markers <- read.csv(f_markers, row.names = NULL)
    markers <- markers[, -1, drop = FALSE]
    colnames(markers) <- c("p", "avg_log2fc", "pct1", "pct2", "p.adj", "cluster", "gene")
    left_join(markers, classes) %>%
        filter(!is.na(cell_type)) %>%
        mutate(signature = paste0("TM:", tissue, ".", cell_type))
}
markers_tmuris <- bind_rows(
    read_tm_markers("Bladder", "facs"),
    read_tm_markers("Spleen", "facs"),
    read_tm_markers("Kidney", "facs"),
    read_tm_markers("Liver", "facs"),
    read_tm_markers("Marrow", "facs"),
    read_tm_markers("Muscle", "facs"),
    read_tm_markers("Lung", "facs"),
    read_tm_markers("Brain_Non-Myeloid", "facs")
)

markers_tmuris <- markers_tmuris %>%
    filter(p.adj <= thresh_padj, avg_log2fc >= thresh_log2fc, pct1 >= thresh_pct1)

markers_tmuris <- markers_tmuris %>%
    group_by(signature) %>%
    summarize(gene = list(gene)) %>%
    deframe()
saveRDS(markers_tmuris, file.path(ROOT, "pipeline", "markers_tmuris.rds"))
```


And MSigDB signatures.

```{r msigdb}
msigdb <- jsonlite::read_json(file.path(MARKERS_DIR, "msigdb_m8.all.v2023.2.Mm.json"))
msigdb <- lapply(msigdb, function(x) unique(unlist(x$geneSymbols)))
names(msigdb) <- paste("msigdb", names(msigdb), sep = ":")
# omit Tabula Muris Senis (ageing) signatures
msigdb <- msigdb[!grepl("TABULA_MURIS_SENIS", names(msigdb))]
saveRDS(msigdb, file.path(ROOT, "pipeline", "markers_msigdb.rds"))
```
