---
title: Mimetic signatures in Tabula Muris Lung Tissue
date: "`r Sys.Date()`"
---

Investigate scoring and assignment of mimetics signatures in Tabula Muris Lung
Tissue.


```{r packages, message = FALSE}
library(scran)
library(scuttle)
library(scater)
devtools::load_all(file.path(ROOT, "R"))
```


```{r}
cts <- read.table(
    file.path(ROOT, "data", "tabula_muris", "FACS", "Lung-counts.csv"),
    row.names = 1, header = TRUE,
    sep = ","
)
meta <- read.csv(
    file.path(ROOT, "data", "tabula_muris", "annotations_facs.csv"),
)
meta <- meta[meta$cell %in% colnames(cts), ]
cts <- as.matrix(cts)
# Parse metadata from identifiers
meta <- data.frame(cell = colnames(cts)) %>%
    separate_wider_delim(
        cell,
        delim = ".",
        names = c("well", "plate.barcode", "mouse.id", NA, NA),
        cols_remove = FALSE
    )
# Get metadata from T. muris
meta_tm <- read.csv(
    file.path(ROOT, "data", "tabula_muris", "annotations_facs.csv"),
)
# Combine
meta <- left_join(meta, meta_tm)

sce_tm <- SingleCellExperiment(assays = list(counts = cts), colData = meta)
```

```{r metadata}
```

```{r}
qcstats <- perCellQCMetrics(sce_tm)
qcfilter <- quickPerCellQC(qcstats)
sce_tm <- sce_tm[, !qcfilter$discard]
clusters <- quickCluster(sce_tm)
sce_tm <- computeSumFactors(sce_tm, clusters = clusters)
summary(sizeFactors(sce_tm))
sce_tm <- logNormCounts(sce_tm)

plot(
    librarySizeFactors(sce_tm), sizeFactors(sce_tm),
    xlab = "Library size factors", ylab = "Deconvolution factors",
    col = rainbow(length(unique(sce_tm$plate.barcode)))[as.integer(as.factor(sce_tm$plate.barcode))],
    log = "xy")

dec <- modelGeneVar(sce_tm)
plot(dec$mean, dec$total, xlab="Mean log-expression", ylab="Variance")
curve(metadata(dec)$trend(x), col="blue", add=TRUE)
chosen.hvgs <- getTopHVGs(sce_tm, prop = 0.1)
sce_tm <- runPCA(sce_tm, subset_row=chosen.hvgs)
sce_tm <- runUMAP(sce_tm, dimred = "PCA", min_dist = 0.3)
```

# Comparative AUCell

Run AUCell on the Tabula Muris Lung data and the TEC data, using the mimetic
Ciliated signature and the Ciliated signature from TM.

```{r}
sce <- readRDS(file.path(ROOT, "pipeline", "sce_integrated.rds"))
# For TECs: get assigned signature
sce_assigned <- readRDS(file.path(ROOT, "pipeline", "sce_integrated_mimetics.rds"))
m <- match(colnames(sce), colnames(sce_assigned))
sce$final_signature <- sce_assigned$signature[m]
sce$source <- "TECs"
# Harmonize column name

# For TM: get original cluster label
sce_tm$source <- "TM_FACS"
sce_tm$final_signature <- sce_tm$cell_ontology_class
sce_tm$transcriptome <- sce_tm$cell

# Restrict to common gene universe
universe <- intersect(rownames(sce), rownames(sce_tm))
sce <- sce[universe, ]
sce_tm <- sce_tm[universe, ]

markers <- readRDS(file.path(ROOT, "pipeline", "markers.rds"))
markers_tmuris <- readRDS(file.path(ROOT, "pipeline", "markers_tmuris.rds"))
eq_sets <- c(
    "Ciliated" = "TM:Lung.ciliated columnar cell of tracheobronchial tree"
)
sets <- c(
    markers[names(eq_sets)],
    markers_tmuris[eq_sets]
)
```

The gene sets are not the same size:

```{r}
lengths(sets)
```

Downsample the larger signature multiple times randomly to the smaller size to
get a range of possible AUC values.

```{r}
set.seed(1)
sets <- lapply(seq_along(eq_sets), function(i) {
    set <- c(sets[eq_sets[i]], sets[names(eq_sets)[i]])
    larger <- which.max(lengths(set))
    smaller <- which.min(lengths(set))
    n_rep <- 100
    larger_downsamp <- replicate(n_rep, {sample(set[[larger]], min(lengths(set)))}, simplify = FALSE)
    names(larger_downsamp) <- paste(names(set)[larger], seq_len(n_rep), sep = "::")
    c(set[smaller], larger_downsamp)
})
sets <- unlist(sets, recursive = FALSE)
```

```{r}
sces <- list(
    tm = sce_tm,
    tecs = sce
)
auc_tbls <- lapply(sces, function(sce) {
    auc <- mimetics::run_aucell(
            sce,
            sets,
            mimetics::AUCELL_MAX_RANK,
            thresholds = list()  # no assignment necessary
    )
    tbl <- auc$tbl
    tbl
})
dat <- bind_rows(auc_tbls) %>%
    separate_wider_delim(signature, "::", names = c("signature", "rep"), too_few = "align_start") %>%
    mutate(rep = ifelse(is.na(rep), 1, rep)) %>%
    group_by(transcriptome, final_signature, source, signature) %>%
    summarize(auc_sd = sd(auc), auc = mean(auc))

```

```{r}
plt_scatter <- function(dat, x, y) {
    dat <- filter(dat, signature %in% c(x, y))
    dat[dat[["signature"]] == x, "signature"] <- "x"
    dat[dat[["signature"]] == y, "signature"] <- "y"
    dat <- dat %>%
        pivot_wider(
            id_cols = c(transcriptome, final_signature, source),
            names_from = signature,
            values_from = c(auc, auc_sd)
        )
    dat[["name_x"]] <- x
    dat[["name_y"]] <- y
    dat
    ggplot(dat, aes(auc_x, auc_y, color = source)) +
        geom_point() +
        geom_errorbar(aes(ymin = auc_y - auc_sd_y, ymax = auc_y + auc_sd_y)) +
        facet_wrap(~final_signature, ncol = 9) +
        geom_abline(slope = 1) +
        coord_fixed() +
        labs(x = x, y = y)
}
sc <- plt_scatter(dat, "Ciliated", "TM:Lung.ciliated columnar cell of tracheobronchial tree")
sc
```
